Field-informed Reinforcement Learning for Learning Large Scale Collective Tasks

### 2023-05-05 paper.tex
feat: add system defintion and problem formalisation. MARL section need to be shrinked

\usepackage[T1]{fontenc}
\usepackage{xcolor}
\usepackage{acronym}
\usepackage{amsmath}
\def\doi#1{\href{https://doi.org/\detokenize{#1}}{\url{https://doi.org/\detokenize{#1}}}}
%
\usepackage{graphicx}

@@

%%% Comment command, to be removed before submission
\newcommand{\meta}[3]{\textcolor{#1}{\textbf{#2}: #3}}
\newcommand{\ga}[1]{\meta{red}{GA}{#1}}
\newcommand{\lukas}[1]{\meta{purple}{Lukas}{#1}}
\newcommand{\mv}[1]{\meta{green}{MV}{#1}}
%% ACRONYMS
\acrodef{DecPOMDP}[DecPOMDP]{decentralized partially observable Markov decision processes}
\acrodef{RL}[RL]{reinforcement learning}
\acrodef{MARL}[MARL]{multi-agent reinforcement learning}
\acrodef{MAARL}[ManyRL]{many-agent reinforcement learning}

\acrodef{MLP}[MLP]{multi-layer perceptron}
\acrodef{GNN}[GNN]{graph neural network}
\acrodef{CNN}[CNN]{convolutional neural network}
\ga{Page limit: 10 pages (include refs)}

@@

\ga{Page budget: 1/2 page \\}
\ga{Plan: discussion about the problem of coordination in multi-agent systems 
and the need for a scalable approach. In doing this, we will discuss some of the existing approaches (declarative, RL and )}
\subsection{Problem definition}
This article focuses on the definition of decentralized collective intelligence within the context of large-scale distributed systems. 
%
Specifically, we consider the case where agents can interact through 1-hop neighbourhood connections and have a local perception of the environment, 
 which is typical in systems such as swarm robotics or large-scale IoT.

We model the system at a given time instance as a graph, $G = (V, E)$, 
 where $E$ defines the neighbourhood relations for each participating node, 
 and $V$ identifies the nodes present in the graph at that time. 
% 
Each node $v$ is associated with an observation (or feature set), $o_v$, 
 which can come from sensors installed on it and messages received from the neighbourhood in the previous step.

Our goal is to find a homogeneously distributed controller $\pi$ that, 
 starting from only local configurations, leads 
 the system to achieve a graph $G_T$ that satisfies certain collective requirements through cooperation, such as spatial area coverage, phenomena tracking, and robot aggregation.
\subsection{Field Coordination Approaches}
\ga{discussion about field used as a coordination mechanism in multi-agent systems. Introduction to AC}
\subsection{Graph Neural Networks}
Over the years, several different neural networks have been proposed to solve specific tasks, ranging from 
  simple \ac{MLP} to \ac{CNN}~\cite{oshea2015introduction}.
%
Whereas \ac{CNN} is designed to be used with image and spatial-like data (e.g., audio),
 \ac{GNN} is a novel model used to process graph-structured data. 
%
Let $ G $ be the above-defined graph.
%
Given $o_v$ a feature vector associated with each node $v \in V$, 
 the goal of a GNN is to learn the node embedding $h_v$ for each node $v \in V$.
 In modern GNNs, the node embedding $h_v$ is computed by aggregating information from the node's neighbours $N(v)$,
  and then combining it with the node's current embedding $h_v$.
% 
Formally, a GNN can be defined by two phases:
\begin{equation}
a_{v}^{(k)}=\operatorname{AGGREGATE}^{(k)}\left(\left\{h_{u}^{(k-1)}: u \in \mathcal{N}(v)\right\}\right)   
\end{equation}
\begin{equation}
h_{v}^{(k)}=\operatorname{COMBINE}^{(k)}\left(h_{v}^{(k-1)}, a_{v}^{(k)}\right)
\end{equation}
where $h_{v}^{k}$ is the embedding of node $v$ at the $k$-th layer ($h_v^0 = o_v$), 
 $\mathcal{N}(v)$ is the set of neighbors of node $v$ computed from $E$.
%
The differential part come into play in the $\operatorname{COMBINE}$ function, 
 which is usually a differentiable function such as a neural network layer.
%
$\operatorname{AGGREGATE} $ instead is a function that aggregates the information from the neighbours of a node $v$
  and it could be a simple sum or a more complex function such as a neural network layer.
%
This formulation allows GNNs to effectively process and extract features from graph-structured data by iteratively aggregating and transforming information from the node's neighbours.

@@

In fact, the first works in this direction were discussed precisely in the last field, 
 exploring new models (e.g., swarMDP~\cite{DBLP:conf/atal/SosicKZK17}) and learning algorithms capable of extrapolating 
 a policy equal to the entire system. 
Modern approaches, however, 
 have started to consider the use of deep learning approaches to try 
 to synthesize robust controllers capable of generalizing to new tasks. 
In this context, mean-field reinforcement learning~\cite{pmlr-v80-yang18d} is certainly noteworthy. 
 Mean-field RL is a technique for learning approximate inference in large-scale systems, 
 where the interactions between the agents can be decoupled into a single conditional probability distribution function for each agent, 
 thus simplifying the learning process and reducing the computational cost.
%
Some known approaches using mean-field reinforcement learning include Q-mean, 
 which is an extension of Q-learning to mean-field settings, 
 and actor-critic mean-field, which combines actor-critic algorithms with mean-field approximations. 
%
These approaches have shown promising results in various domains, such as multi-agent coordination 
 and decentralized control, and are actively being researched and developed for further applications

\section{Field-informed Reinforcement Learning}
\label{sec:approach}
\ga{Page budget: 2/3 pages \\}
\ga{Plan: we should discuss the approach in detail, including the system model, and the learning dynamics.
Particularly, I would like to highlight how each node in the graph is a local controller,
but it could be seen as a global evolution, therefore we can use global information to inform the local controller.}
\subsection{System model}
\ga{Plan: 
  discussion about the typical AC evaluation model, specifying how GNN (with 1-hop information diffusion)
  can be easily integrated with this distributed model
}
\subsection{Learning dynamics}
\ga{
  Plan: discuss the chosen approach, that is Centralised Training and Distributed Execution.
  Particularly, we use DQN in which the neural network is a GNN. So, the dataset consists in a set of graphs,
  where each graph is a snapshot of the system. 
  The target is the Q-value of the action taken in the current state.
  Explain how this can be then used for distributed controllers.
}
\section{Evaluation}
\label{sec:eval}
\ga{Page budget: 2 pages \\}
\ga{Plan: we should discuss the robot aggregation scenario, specifying the state space, the action space, and the reward function.
I dunno if it makes sense to discuss the variants (i.e., fixed area, two areas and moving area). We will see.
}
\subsection{Scenario}

>>>>

\usepackage[T1]{fontenc}
\usepackage{amsfonts}
\usepackage{xcolor}
\usepackage{acronym}
\usepackage{amsmath}
\def\doi#1{\href{https://doi.org/\detokenize{#1}}{\url{https://doi.org/\detokenize{#1}}}}
%
\usepackage{graphicx}

@@

%%% Comment command, to be removed before submission
\newcommand{\meta}[3]{\textcolor{#1}{\textbf{#2}: #3}}
\newcommand{\ga}[1]{\meta{red}{GA}{#1}}
\newcommand{\lukas}[1]{\meta{purple}{Lukas}{#1}}
\newcommand{\mv}[1]{\meta{green}{MV}{#1}}
%% ACRONYMS
\acrodef{DecPOMDP}[DecPOMDP]{decentralized partially observable Markov decision processes}
\acrodef{RL}[RL]{reinforcement learning}
\acrodef{MARL}[MARL]{multi-agent reinforcement learning}
\acrodef{MAARL}[ManyRL]{many-agent reinforcement learning}
\acrodef{MDP}[MDP]{Markov Decision Process}
\acrodef{MLP}[MLP]{multi-layer perceptron}
\acrodef{GNN}[GNN]{graph neural network}
\acrodef{CNN}[CNN]{convolutional neural network}
\ga{Page limit: 10 pages (include refs)}

@@

\ga{Page budget: 1/2 page \\}
\ga{Plan: discussion about the problem of coordination in multi-agent systems 
and the need for a scalable approach. In doing this, we will discuss some of the existing approaches (declarative, RL and )}
\subsection{Problem definition}
This article focuses on the definition of \emph{decentralized collective intelligence} within the context of \emph{large-scale} distributed systems. 
%
Specifically, we consider systems of \emph{devices} (or nodes) equipped with \emph{sensors} and \emph{actuator}
 each of which can only interact with the \emph{1-hop} neighbourhood that can 
 be logical (i.e., a communication network) or physical (i.e., a spatial neighbourhood).
%
Each node executes a \emph{local} controller that can only access the information available to its \emph{context}
 that is the information available in the neighbourhood and the information available from its sensors.
%
After each program evaluation, 
 the node can send messages to its neighbourhood about its current evaluation and update its internal state.
%
This general model can be adapted to several different scenarios, 
 such as a network of sensors, a network of robots (e.g., swarm robotics), or a network of mobile phones.
%
Our goal is to find a \emph{homogeneously} distributed controller $\pi$ (i.e., the same controller for the whole system) that, 
 starting from \emph{only} local configurations, leads 
 the system to achieve a certain \emph{collective} requirements through \emph{cooperation}, 
 such as spatial area coverage, phenomena tracking, and robot aggregation.
%
The homogeneity of the controller is a key requirement to ensure the \emph{scalability} of the approach, 
 as it allows us to avoid the need for a centralised controller that would be a bottleneck for the system,
 and it is the typical choice in swarm-like systems~\cite{brambilla2013swarm,yang2021many,pmlr-v80-yang18d,DBLP:conf/aaai/ZhengYCZZWY18}.
%
\subsection{Field Coordination Approaches}
\ga{discussion about field used as a coordination mechanism in multi-agent systems. Introduction to AC}
\subsection{Graph Neural Networks}
Over the years, several different neural networks have been proposed to solve specific tasks, ranging from 
  simple \ac{MLP} to \ac{CNN}~\cite{oshea2015introduction}.
%
Whereas \ac{CNN} is designed to be used with image and spatial-like data (e.g., audio),
 \ac{GNN} is a novel model used to process graph-structured data. 
%
Let $G = (V, E)$ a graph  
where $E$ defines the neighbourhood relations for each participating node, 
and $V$ identifies the nodes present in the graph at that time. 
% 
Each node $v$ is associated with an observation (or feature set), $o_v$, 
%
Given $o_v$ a feature vector associated with each node $v \in V$, 
 the goal of a \ac{GNN} is to learn the node embedding $h_v$ for each node $v \in V$.
 In modern \acp{GNN}, the node embedding $h_v$ is computed by aggregating information from the node's neighbours $N(v)$,
  and then combining it with the node's current embedding $h_v$.
% 
Formally, a \ac{GNN} can be defined by two phases:
\begin{equation}
a_{v}^{(k)}=\operatorname{AGGREGATE}^{(k)}\left(\left\{h_{u}^{(k-1)}: u \in \mathcal{N}(v)\right\}\right)   
\end{equation}
\begin{equation}
h_{v}^{(k)}=\operatorname{COMBINE}^{(k)}\left(h_{v}^{(k-1)}, a_{v}^{(k)}\right)
\end{equation}
where $h_{v}^{k}$ is the embedding of node $v$ at the $k$-th layer ($h_v^0 = o_v$), 
 $\mathcal{N}(v)$ is the set of neighbors of node $v$ computed from $E$.
%
The differential part come into play in the $\operatorname{COMBINE}$ function, 
 which is usually a differentiable function such as a neural network layer.
%
$\operatorname{AGGREGATE} $ instead is a function that aggregates the information from the neighbours of a node $v$
  and it could be a simple sum or a more complex function such as a neural network layer.
%
This formulation allows GNNs to effectively process and extract features from graph-structured data by iteratively aggregating and transforming information from the node's neighbours.

@@

In fact, the first works in this direction were discussed precisely in the last field, 
 exploring new models (e.g., swarMDP~\cite{DBLP:conf/atal/SosicKZK17}) and learning algorithms capable of extrapolating 
 a policy equal to the entire system. 
Modern approaches, however, 
 have started to consider the use of deep learning approaches to try 
 to synthesize robust controllers capable of generalizing to new tasks. 
In this context, mean-field reinforcement learning~\cite{pmlr-v80-yang18d} is certainly noteworthy. 
 Mean-field RL is a technique for learning approximate inference in large-scale systems, 
 where the interactions between the agents can be decoupled into a single conditional probability distribution function for each agent, 
 thus simplifying the learning process and reducing the computational cost.
%
Some known approaches using mean-field reinforcement learning include Q-mean, 
 which is an extension of Q-learning to mean-field settings, 
 and actor-critic mean-field, which combines actor-critic algorithms with mean-field approximations. 
%
These approaches have shown promising results in various domains, such as multi-agent coordination 
 and decentralized control, and are actively being researched and developed for further applications
\subsection{Problem formalisation}
Given the homogeneity and large scale considered and the \emph{locality} (i,e., each agent can only observe its neighbours), 
 the problem can be modelled through the SwarMDP model~\cite{DBLP:conf/atal/SosicKZK17} -- 
 an extension of the \ac{DecPOMDP}~\cite{Decpomdp2000} model to swarm-like systems. \ga{todo: define swarm-like systems}
% 
A SwarMDP is characterized by a \emph{swarming agent} ($\mathbb{A}$) and the dynamics of the environment ($\mathbb{E}$).
Specifically, $\mathbb{A}$ is a tuple ($\mathcal{S}, \mathcal{O}, \mathcal{A}, \mathcal{R}, \pi$) where:
\begin{itemize}
  \item $\mathcal{S, O, A}$ are the set of local states, observations (or features), and actions, respectively;
  \item $\mathcal{R}: \mathcal{S} \rightarrow \mathbb{R}$ is the reward function, which is influenced by the environment;
  \item $\pi: \mathcal{O} \rightarrow \mathcal{A}$ is the policy function, which maps the observations to the actions. It could be deterministic or stochastic.
\end{itemize}
Starting from this definition, the environment $\mathbb{E}$ is defined as a tuple ($\mathcal{P}, \mathbb{A}, \mathcal{T}, \xi$), where:
\begin{itemize}
  \item $\mathcal{P}$ is the total number of agents in the systems (the agent population), which is assumed to be fixed;
  \item $\mathbb{A}$ is the above defined agent prototype that rules each agent $v \in P$;
  \item $\mathcal{T}: \mathcal{S}^P \times \mathcal{A}^P \times \mathcal{S}^P \rightarrow \mathbb{R}$ is the transition  global function, which is influenced by the actions of the agents and returns a collective reward -- this is typically not known by the swarming agents;
  \item $\xi: \mathcal{S^P} \rightarrow \mathcal{O^P}$ is the global observation model of the systems.
\end{itemize}
In swarMDP, the neighbourhood is not directly defined, but it is implicitly defined by the observation model $\xi$.
Since in our case, the agents can only interact with 1-hop neighbourhoods and they aren't directly influenced by other 
  agent observations, we restrict the observation model as follows:
\begin{equation}
\xi(v): \{s_j, j \in \mathcal{N}^v\} \rightarrow O
\end{equation}
\begin{equation}
\xi = \{\xi(v), v \in \mathcal{P}\}
\end{equation}
where $\mathcal{N}^v$ is the set of neighbours of $v$.

This model can be used then to express the evolution of the system in time.
Specifically, starting from a global state $\mathcal{S}^P_t$, the next state $\mathcal{S}^P_{t+1}$ is defined as:
\begin{equation}
\mathcal{A}^P_t = \pi(\xi(S^P_t))
\end{equation}
\begin{equation}
\mathcal{S}^P_{t+1} = \mathcal{T}(\mathcal{S}^P_t, \mathcal{A}^P_t)
\end{equation}
Given a time $t$, the system can be also represented as a graph $G_t = (V_t, E_t)$, 
  where $E_t$ is built from $\mathcal{N}$.
  Each node is then decorated with the local observation $o_v$. 
  Therefore, this graph can be used for being processed by a \ac{GNN} as done in~\cite{DBLP:conf/corl/TolstayaGPP0R19,tolstaya2020learning,DBLP:conf/icra/GosrichMLPYR022}.
\section{Field-informed Reinforcement Learning}
\label{sec:approach}
\ga{Page budget: 2/3 pages \\}
\ga{Plan: we should discuss the approach in detail, including the system model, and the learning dynamics.
Particularly, I would like to highlight how each node in the graph is a local controller,
but it could be seen as a global evolution, therefore we can use global information to inform the local controller.}
\subsection{System model}
\ga{Plan: 
  discussion about the typical AC evaluation model, specifying how GNN (with 1-hop information diffusion)
  can be easily integrated with this distributed model
}
\subsection{Learning dynamics}
\ga{
  Plan: discuss the chosen approach, that is Centralised Training and Distributed Execution.
  Particularly, we use DQN in which the neural network is a GNN. So, the dataset consists in a set of graphs,
  where each graph is a snapshot of the system. 
  The target is the Q-value of the action taken in the current state.
  Explain how this can be then used for distributed controllers.
}
\section{Evaluation}
\label{sec:eval}
\ga{Page budget: 2 pages \\}
\ga{Plan: we should discuss the robot aggregation scenario, specifying the state space, the action space, and the reward function.
I dunno if it makes sense to discuss the variants (i.e., fixed area, two areas and moving area). We will see.
}
\subsection{Scenario}
